We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-27 paper/paper.tex
edit conclusions

In this paper we develop and demostrate a method to quantify the
asphericity of the satellite distribution in the around the MW and M31.  
In the interest of keeping the method straightforward and robust we do
not look for planes in subsets of satellites or coherent kinematical
structures. 
We focus on the asphericity estimates for a fixed number of the
brightest satellites around each galaxy.

The method uses as a reference the spherically randomized data
of the system under study \citep{2017AN....338..854P}.  
To this end, we first measure the width and axis ratios for the satellite
distributions of interest. 
Then, we measure the same quantities for the same set of points after
a process of spherical randomization.
Finally, we renormalize the initial results to the mean value and
standard deviation computed from the randomized data.

We find that these normalized quantities are well described by
a multivariate gaussian distribution. 
We fit estimate the mean and covariance of this distributions using
the results of LG pairs coming from three different numerical
simulations, we inally compare the observational results against the
distributions derived from the simulations: Illustris-1,
Illustris-1-Dark and ELVIS.

We find that in the best case (Illustris1-DarkMatter) the deviation
from asphericity in the observed LG is only expected in $3\pm2$
pairs out of a sample of $10^4$ isolated pairs. This places the LG as
a $3\sigma$ outlier.   
The weight to explain this atypical result is not distributed equally
between the MW and M31. 
While M31 presents a fully typical asphericity in the expectations
from LCDM, the MW shows aspherical deviations in plane width and the
major-to-minor axis ratio highly atypical in the 
framework of LCDM. 
We estimate that with the M31 between $37\%$ and $58\%$ of the pairs
pairs show normalized characteristics larger than M31, while
this fraction drops to less than $1\%$ for the MW.
These fractions are robust to changes in the numerical simulations and to
the criteria used to define the pairs.

The highly aspherical distribution of the brightest MW satellites is
in stark contrast to the more spherical M31 distribution.
Furthermore, the atipicallity of the MW distribution in the LCDM
context adds up to the also atypical presence of two bright satellite
galaxies.

The focus of our approach is building explicit probability
distributions for the observables of interest, instead of trying to
find simulated objects that fulfill different observational criteria. 
This approach is particularly useful in the case of atypical
observables. 
Building the probability distribution allows for an atypicallity
quantification without needing to build explicit samples of objects
that are already scarce and difficult to find in simulations.

An extension of this framework to outliers in higher order deviations
(i.e. a \emph{subset} of satellites that are in a plane or
coherent \emph{velocity} structures) should also be possible, similar
to the work presented by XX\cite{2013Natur.493...62I} on the satellite velocity anisotropy.

From the results we derive in this paper we also estimate that a
cosmological volume of at least $( XX )^3$ is required to find a halo pair
that shows similar asphericity characteristics as the LG. 
This atypical satellite distribution should be seen as an opportunity
to constrain in high detail the initial conditions and environment
that allowed such a pattern to emerge.

>>>>

In this paper we developed and demostrated a method to quantify the
asphericity of the satellite distribution around the MW and M31.  
In the interest of keeping the method straightforward and robust we
focus on the asphericity estimates for a fixed number of the brightest
satellites around each galaxy.

The method uses as a reference the spherically randomized data
of the system under study \citep{2017AN....338..854P}.  
To this end, we first measure the width and axis ratios for the satellite
distributions of interest. 
Then, we measure the same quantities for the same set of points after
the spherical randomziation process.
Finally, we renormalize the initial results to the mean value and
standard deviation computed from the randomized data.

We found that these normalized quantities are well described by
a multivariate gaussian distribution in spite of the original
quantities being bound.
We estimated the mean and covariance of this distributions using
the results of LG pairs coming from three different numerical
simulations (Illustris-1, Illustris-1-Dark and ELVIS). 
Finally, we compared the observational results against the
distributions derived from the simulations

We found that in the best case (Illustris1-Dark) the degree of
asphericity in the observed LG is only expected in $3\pm2$  pairs out
of a sample of $10^4$ isolated pairs. 
This places the LG as a $3\sigma$ outlier.   
The weight to explain this atypical result is not distributed equally
between the MW and M31. 
While M31 presents a fully typical asphericity in the expectations
from LCDM, the MW shows aspherical deviations in plane width and the
major-to-minor axis ratio highly atypical in the  framework of LCDM,
confirming the original hint by \cite{2005A&A...431..517K} and more
recent results by \cite{2015ApJ...815...19P}. 
We estimated that with the M31 between $37\%$ and $58\%$ of the pairs
pairs show aspherical characteristics larger than M31, while
this fraction drops to less than $1\%$ for the MW. 
These fractions are robust to changes in the numerical simulations
and, the criteria used to define the pairs and the methods used to
estimate the parameters of the multivariate normal distribution.

The focus of our approach was building an explicit probability
distributions for the observables of interest, instead of trying to
find simulated objects that fulfill different observational criteria. 
This approach is particularly useful in the case of atypical
observables as its the case of interest, as it allows for an atypicallity
quantification without needing to build explicit samples of objects
that are already scarce and difficult to find in simulations.
An extension of this framework to outliers in higher order deviations
(i.e. coherent \emph{velocity} structures) should also be possible, 
provided that an explicit probability distribution for the scalars of
interest can be built.

The highly aspherical satellite distribution in the MW is another piece of
information that points at an atypical configuration in LCDM.
We also have the number of satellites as bright as the Mageallanic
Clouds, only expected in $5\%$ of galaxies \citep{2011ApJ...743..117B}
and the atypical kinematics of M31 with a very low tangential velocity,
only expected in less than $1\%$ of the pairs with similar
environmental characteristics \citep{ForeroRomero2013}. 
In other words, if we were in a typical LG pair in LCDM, the MW should
have zero satellites like the Magellanic Clouds, it's satellite
distribution should be similar to M31's and M31 should have a larger
tangential velocity.

The LG atypicality should be seen as an opportunity to constrain in
high detail the environment that allowed such a pattern to emerge. 
Although local correlations between LG assembly, pair kinematics, halo shapes and
satellite distributions are expected in LCDM
\citep{2011MNRAS.417.1434F,2014MNRAS.443.1090F,2015ApJ...799...45F,2015MNRAS.452.1052L},
we claim that detailed studies of the satellite asphericity as a function
of halo mass and cosmic web environment are still missing to
understand what fearures in the initial conditions of our Local Group
are responsible for the assembly extreme features observed in its satellite
distribution.
